The curious world of constructive mathematics

### 2022-09-13 algar-lecture-notes.tex
More work on KJ examples and the generic prime filter

\newcommand{\cl}{\mathrm{cl}}
\newcommand{\notnot}{\emph{not~not}}
\newcommand{\Sh}{\mathrm{Sh}}
\newcommand{\Set}{\mathrm{Set}}

@@

\begin{tabular}{@{}lp{0.83\textwidth}ll@{}}
  \toprule
  & statement & status \\\midrule
  1a & Every continuous function~$f : [-1,1] \to \RR$ with~$f(-1) < 0$
  and~$f(1) > 0$ has a zero. & \cmark/\xmark \\
  1b & For every continuous function~$f : X \times [-1,1] \to \RR$
  with~$f(\cdot,-1) < 0$ and~$f(\cdot,1) > 0$ there is an open covering~$X =
  \bigcup_i U_i$ such that for each index~$i$, there is a continuous
  zero-picking function~$z : U_i \to [-1,1]$
\end{tabular}

@@

A rewarding explanation of this phenomenon is nowadays relayed in any course on
complex analysis: The function~$1/(1+x^2)$ does have a singularity at
distance~$1$ from the origin, namely at~$\pm i$ in the complex plane, and the disk
of convergence is only so big that it merely touches the nearest singularity.

This episode is one example of many how the ``imaginary unit''---once
an elusive \emph{mathematical phantom}---affects the story of the real numbers even
though it itself is not one of them. Like many mathematical phantoms before it,
the imaginary unit obtruded its effects so convincingly that eventually we
embraced it and broadened our notion of existence.

@@

\section{Ring-theoretic prelude: the method of indeterminates}

@@

\section{The generic prime filter}
\newpage
\ \newpage

@@

@@

\begin{exercise}[Anonymously Noetherian rings]
A commutative ring~$A$ is \emph{anonymously Noetherian} if for every
ideal~$\aaa \subseteq A$ it is \notnot the case that~$\aaa$ is finitely
generated.
\begin{enumerate}
\item Show that residue fields in the sense of Footnote~\ref{fn:fields} are
anonymously Noetherian.
\item Verify that~$\ZZ$ is anonymously Noetherian.

>>>>

\newcommand{\cl}{\mathrm{cl}}
\newcommand{\notnot}{\emph{not~not}\xspace}
\newcommand{\Sh}{\mathrm{Sh}}
\newcommand{\Set}{\mathrm{Set}}

@@

\emph{Analysis:}
\begin{enumerate}
\item[\cmark/\xmark] Every continuous function~$f : [-1,1] \to \RR$ with~$f(-1) < 0$
  and~$f(1) > 0$ has a zero.
\item[\xmark/\xmark] Let~$X$ be a topological space and let~$f : X \times [-1,1] \to \RR$
be a continuous function with~$f(\cdot,-1) < 0$ and~$f(\cdot,1) > 0$. Then
there is an open covering~$X = \bigcup_i U_i$ such that for each index~$i$,
there is a continuous zero-picking function~$z : U_i \to [-1,1]$.
\end{enumerate}

\emph{Linear algebra:}
\begin{enumerate}
\item[\cmark/\xmark] Every real symmetric matrix has an eigenvector.
\item[\xmark/\xmark] Let~$X$ be a topological space and let~$A : X \to
M^\text{sym}_n(\RR)$ be a continuous map to the space of symmetric~$(n \times
n)$-matrices. Then there is an open covering~$X = \bigcup_{i \in I} U_i$ such
that or all indices~$i \in I$, there is a continuous map~$v : U_i \to \RR^n$
such that for each~$x \in U_i$, the vector~$v(x)$ is an eigenvector of~$A(x)$.
\end{enumerate}

\emph{Commutative algebra:}
\begin{enumerate}
\item[\cmark/\cmark] Let~$M$ be a finitely generated projective module over a local
ring~$A$. Then~$M$ is finite free.
\item[\cmark/\cmark] Let~$M$ be a finitely generated projective module over an
arbitrary commutative ring~$A$. Then there is a partition~$1 = f_1 + \cdots +
f_n \in A$ of unity such that, for each index~$i$, the localized
module~$M[f_i^{-1}]$ is finite free over~$A[f_i^{-1}]$.
\item[\cmark/\xmark] Let~$M$ be a finitely generated module over a (residue)
field~$k$. Then~$M$ is finite free.
\item[\xmark/\xmark] Let~$M$ be a finitely generated module over an
arbitrary commutative ring~$A$. Then there is a partition~$1 = f_1 + \cdots +
f_n \in A$ of unity such that, for each index~$i$, the localized
module~$M[f_i^{-1}]$ is finite free over~$A[f_i^{-1}]$.
\item[\cmark/\cmark] Let~$M$ be a finitely generated module over a (residue)
field~$k$. Then~$M$ is \notnot finite free.
\item[\cmark/\cmark] Let~$M$ be a finitely generated module over an
arbitrary commutative ring~$A$. If~$f = 0$ is the only element of~$A$ such
that~$M[f^{-1}]$ is finite free over~$A[f^{-1}]$, then~$1 = 0$ in~$A$.
\item[\cmark/\cmark] If~$k$ is a (residue) field, there is no linear
surjection~$k^n \to k^m$ with~$m > n$.
\item[\cmark/\cmark] Let~$A$ be an arbitrary commutative ring. If there exists
a linear surjection~$A^n \to A^m$ with~$m > n$, then~$1 = 0$ in~$A$.
\end{enumerate}

@@

A rewarding explanation of this phenomenon is nowadays relayed in any course on
complex analysis: The function~$1/(1+x^2)$ does have a singularity at
distance~$1$ from the origin, namely at~$\pm i$ in the complex plane, and the disk
of convergence is and can only be so big that it just touches the nearest singularity.

This episode is one example of many how the ``imaginary unit''---once
an elusive \emph{mathematical phantom}---affects the story of the real numbers even
though it itself is not one of them. Like many mathematical phantoms before it,
the imaginary unit obtruded its effects so convincingly that eventually we
embraced it and broadened our notion of existence~\cite{wraith:phantoms}.

@@

\section{Ring-theoretic prelude: the method of indeterminates}
\label{sect:indeterminates}

@@

\section{The generic prime filter}

A commutative ring~$A$ is often studied by its \emph{stalks}~$A_\ppp$ at prime
filters~$\ppp \subseteq A$, those subsets of~$A$ which validate the conditions
on the right:
\begin{align*}
  \top &\,\,\Longrightarrow\, 0 \in \ppp &
  0 \in \ppp &\,\,\Longrightarrow\, \bot \\
  x \in \ppp \wedge y \in \ppp &\,\,\Longrightarrow\, x + y \in \ppp &
  x + y \in \ppp &\,\,\Longrightarrow\, x \in \ppp \vee y \in \ppp \\
  1 \in \ppp &\,\,\Longrightarrow\, \bot &
  \top &\,\,\Longrightarrow\, 1 \in\ppp \\
  x \in \ppp \vee y \in \ppp &\,\Longleftrightarrow\, xy \in \ppp &
  xy \in \ppp &\,\Longleftrightarrow\, x \in \ppp \wedge y \in \ppp
\end{align*}
The stalk~$A_\ppp$ is then the localization~$A[\ppp^{-1}]$, that is
the ring of formal fractions~$x/s$ with~$x \in A$ and~$s \in \ppp$, where~$x/s
= y/t$ iff there exists an element~$u \in \ppp$ such that~$utx = usy$.
The set~$\ppp$ is multiplicatively closed (and saturated) by the last two
conditions.\footnote{Textbooks on classical commutative algebra usually use
\emph{prime ideals} instead of \emph{prime filters}, those subsets which
validate the dual conditions on the left. In classical commutative algebra, we
can freely pass between these two notions by taking complements, as the
complement of a prime ideal is a prime filter and conversely. Constructivly,
prime filters are more useful, and even classically they are slightly more
convenient to work with.} Similarly, an~$A$-module~$M$ is often studied by its
stalks~$M_\ppp = M[\ppp^{-1}]$.

Stalks are interesting because they allow us to zoom in on ``local''
situations---they tend to be simpler than the original ring or module, while
still being intimately connected to the originals. For instance:
\begin{enumerate}
\item The stalks~$A_\ppp$ are always \emph{local
rings},\label{page:local-rings} rings such that a finite sum is invertible only
if one of its summands also is. (This amounts to~$1 \neq 0$ and that
whenever~$x+y$ is invertible, $x$ is invertible or~$y$ is invertible; see
Exercise~\ref{ex:local-rings}.)
\item Classically, if~$A$ is reduced, the stalks~$A_\ppp$ at
\emph{maximal} prime filters (equivalently minimal prime ideals) are even fields.
\item Classically, an~$A$-module~$M$ is flat if and only if all of its stalks~$M_\ppp$ are.
\end{enumerate}

@@

\begin{exercise}[Verifying polynomial identities]

\end{exercise}

@@

\begin{exercise}[Local rings]\label{ex:local-rings}
\begin{enumerate}
\item Using the definition of local ring given on
page~\pageref{page:local-rings}, verify that every geometric field in the sense
of Footnote~\ref{fn:fields} is a local ring; that the (Cauchy or Dedekind)
reals form a local ring; that for each prime number~$p$ the subring of~$\QQ$
consisting of those fractions whose denominator is not divisible by~$p$ is a
local ring; and that~$\ZZ$ and~$\QQ[X]$ are not local rings.

{\noindent\scriptsize\emph{Note.} If~$a < b$ are real numbers, then for every
real number~$x$, $x > a$ or~$x < b$.\par}

\item Show that the stalk~$A_\ppp$ of a commutative ring~$A$ at a prime
filter~$\ppp$ is a local ring. What goes wrong in case that~$\ppp$ is instead a
prime ideal and one considers the localization~$A[(A\setminus\ppp)^{-1}]$?

\item Verify in classical mathematics that a ring is local if and only if it
has exactly one maximal ideal.

{\noindent\scriptsize\emph{Hint.} Use that, in classical mathematics, every
nonunit is contained in some maximal ideal.\par}
\end{enumerate}
\end{exercise}

\begin{exercise}[Abstract existence of bounds from the method of
indeterminates]
\begin{enumerate}
\item Using Krull's lemma in its unconstructive form, verify for every
commutative ring~$A$ that if a polynomial~$f \in A[X]$ is nilpotent in~$A[X]$,
all its coefficients are nilpotent in~$A$.
\item Let~$f \in A[X]$ be a polynomial of degree~$d$ such that~$f^n = 0$.
By the proof in~(a), there exists a number~$m$ such that the~$m$-th power of
each of the coefficients of~$f$ vanishes. However, it is conceivable that~$m$
depends not only on~$d$ and~$n$, but also on the ring~$A$ and the specific
values of the coefficients. Explain how the method of indeterminates reviewed
in Section~\ref{sect:indeterminates} can be used to establish that there exists
some bound on~$m$ depending only on~$d$ and~$n$.
\end{enumerate}
\end{exercise}

\begin{exercise}[Anonymously Noetherian rings]
A commutative ring~$A$ is \emph{anonymously Noetherian} iff for every
ideal~$\aaa \subseteq A$ it is \notnot the case that~$\aaa$ is finitely
generated.
\begin{enumerate}
\item Show that residue fields in the sense of Footnote~\ref{fn:fields} are
anonymously Noetherian.
\item Verify that~$\ZZ$ is anonymously Noetherian.
